Towards an Ecological Trait-data Standard

### 2017-11-13 manuscript/manuscript.Rmd
finalize manuscript, spell checked (british), added examples to R package, shorten text on extensions

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider
- name: Malte Jochum
- name: Gaetane LeProvost
- name: Andreas Ostrowski
- name: Caterina Penone
- name: Nadja K. Simons
csl: amnat.csl
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

1.	The assessment and analysis of functional traits has been established in all domains of ecological research, promising great advancements for biodiversity and climate research. A wealth of data is available from multiple sources and for many organism groups but is lacking standardisation due to heterogeneous data formats and data types. 
2.	To harmonize and standardize datasets, we present a general vocabulary and structure for an ecological trait-data standard for storing and sharing functional trait data.
3.	We further provide an R-package that assists in transferring own data structures into the data standard and to harmonize and merge trait datasets from heterogeneous sources. 
4.	With these decentralized tools for traitdata management, we intend to facilitate the exchange and analysis of trait data for ecological research.

@@

Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. 
Mutual 'trait-matching' of co-evolved organisms enables the inferrence of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parameterize ecological population-dynamic models [@schneider12; @kalinkat13]. These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16]. 
Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @jennings09; @simons16; @gossner16; @moretti13; @gamez-virues15; @kuhsel15;  @birkhofer14; @gravel16; @lajoie15], as well as macro-patterns of biodiversity at the global level [@diaz13; @diaz16; @reich99; @reich97]. Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].
Most importantly, the trait framework relates key ecosystem services for human well-being back to changes in biodiversity in consequence to global change or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17].

In recent years, functional ecology has thus grown into a major field of ecological research, and that continues to integrate more and more fundamental and applied research questions. 
Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. 
However, it relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa. 
To mobilize data for this enterprise, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources. The development of semantic-web standards [@wieczorek12; @page08] and ontologies [@walls12; @garnier17], as well as web-services [@kattge11; @parr16] for trait-data management is under way. 
Nonetheless, individual researchers often neglect the benefit of centralised databases and, to cope with journals' and research funding agencies increasing demand for the publication of data without access restrictions [@swan12; @allison15], instead publish primary data on file hosting services, where no data standards are forced upon the uploaded material (e.g. Figshare, DataDryad). Given this prospect, there will be no lack of data, but a lack of data standardisation.

In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from an pragmatic view of data providers and data users. We further (2.) review existing trait databases and web services, as well as initiatives for standardisation. 
We (3.) describe the common practice for structuring trait data for ecological research. 
Based on these considerations, we (4.) propose a low-threshold struture and vocabulary for describing trait-datasets compatible with existing ontologies. (5.) We provide a tool for an easy adoption of this structure and for the harmonisation of data from different sources.

@@

Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12; @ulrich08] derived from metagenomic analysis [@torsvik02; @langille13]. Databases are at hand to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g. the KEGG orthology, @kanehisa12].

@@

A global overview of existing trait data for all taxa and trait types is difficult to obtain. As an attempt to track  databases and unstandardized datasets from indipendent publications, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets with their their taxonomic focus and extend of traits covered, their location on the internet and the terms of use. We invite data owners and users to register further datasets qualifying as trait data.

@@

Lists of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html @hoehndorf16]. On the side of animal traits, a couple of initiatives published comprehensive trait standards, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13]. A resource for biological ontologies and terminologies is the OBO foundry (http://obofoundry.org/).

Relating the trait definition to methodological definitions on trait measurements are key for harmonizing trait data already at the physical process of measureing. For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in Thesauri.

While trait thesauri define traits for focal groups of organisms, they are not specifing the  formats or structures in which trait data should be stored. 
The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables. The framework employs established terms provided by the Darwin Core Standard [DWC, @wieczorek12], relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g. GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr16]. Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. 
Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].

@@

Quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig. 1*a*), by aggregation of replicated measurements on multiple entities (Fig. 1*b*) or by estimation of means or ranges for the species as reported in literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1.*d*). The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a subsample or bodypart, an individual specimen, an entire species or higher-level taxon.  
This definition refines the entity-quality model (EQ) as described by @garnier17, where a trait observation is phrased as 'an entity having a quality'. 
<!-- Cat: the difference [between quantitative measurements or qualitative facts] is not totally clear to me - is it the difference between field vs literature gathered data?  Flo: I see measurements as quantitative and facts as qualitative traits, which mostly aligns with field vs. literature, but not entirely. A differentiation that has rarely been specified in the trait-litereature it seems. Garnier et al 2017 just speak of "quality", which is assigned to an "entity", without specifiying it further. Nadja: I like this differentiation, to me this nicely reflects morphometric and literature traits (although its of course not exclusive). Maybe it helps if we add an example after the definition? --> <!--FDS: is this sufficiently defined now? the distinction between measured and literature data is spelled-out further below, when discussing the formats. Quantitative values can be taken from literature, though, like body sizes for example. -->

\begin{figure}[t]
\centering
\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
\caption{Types of ecological trait data assume different entities or reported quantities. a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g. reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, refering to the entire taxon (or a subset, e.g. a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire taxon.}

@@

Trait-datasets usually take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (a.k.a. wide-table format,  Fig. 2*a*), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as 'NA'. It may store additional information (e.g. on variation of means or literature source) in secondary colums or accompanying datasheets. 
The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon). Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also at this level of information, missing entries, e.g. about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using bayesian approaches [@schrodt15]. However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnick16].

@@

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence). This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.  Longtable dataset purport multiple advantages for the data manipulation (e.g. filtering, subsetting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modeling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. They further draw from a defined set of columns, which simplifies merging of datasets.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
\caption{Formats used for trait datasets. Identifiers relate the measurement to other layers of information on the specimen, the taxon or the trait measured. a) taxon-level traits are often published as a compiled species $\times$ traits matrix; b) measured data report measurements on the specimen-level, i.e. one physical occurrence of a taxon; c) a measurement longtable is a well defined and tidy data format, reporting one single measurement per row and relating it to an occurrence and taxon via unambiguous identifiers.   }

@@

With this paper, we propose a versatile, minimal data structure, as well as a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research. We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major webservices and existing standards for biodiversity data management.

@@

As described above, tidy longtable formats are the most predictable and compatible format for the core content of trait datasets and should be used for standalone uploads on low-threshold repositories.

To enable a better standardization of nomenclatures and values, the entries of the core dataset should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names. In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08]. 
If data contain multilayered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below). 
For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt').

@@

In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14]. Therefore we recommend use of  Darwin Core Archives [DwCA; @robertson09], the structure proposed for GBIF [@gbif17; http://tools.gbif.org/dwca-assistant/] or EOL TraitBank [@parr15; http://eol.org/info/cp_archives]. This also integrates a defined framework for metadata in the Ecological Metadata Language [@knb11], which should be used for describing and linking the data tables.

@@

Column names and, where possible, field entries should follow the defined vocabulary listed at https://ecologicaltraitdata.github.io/TraitDataStandard/. 
Individual terms should be referenced via their global identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e. via the Application Programming Interface, API). Whereever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

@@

It is indispensible to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits. This is acheived by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively), according to a public ontology (e.g. GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass').  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practiced in TRY; @kattge11]. This redundancy of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. Keeping an additional reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
It is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].

The existing thesauri and ontologies differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. More compatibility of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.

If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. 
We encourage publishing trait thesauri developed for the own project context as public resources, e.g. on OBO Foundry, the GFBio Terminology service, or on other stable web services.

See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/TraitDataStandard/#terms-for-traitlists-a-trait-thesaurus].  We also provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/). 
### Metadata

Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
We define a set of terms that allows to relate trait data to authorship and define a bibliographic reference and license model.

@@

<!--
### additional information on measurement or fact

As a high-level discrimination of the source of the measurement or fact, the Darwin Core Term `basisOfRecord` takes an entry about the type of trait data recorded: Were they taken by own measurement (distinguish "LivingSpecimen", "PreservedSpecimen", "FossilSpecimen") or taken from literature ("literatureData"), from an existing trait database ("traitDatabase"), or is it expert knowledge ("expertKnowledge"). It is highly recommended to provide further detail about the source in the column `basisOfRecordDescription`.

>>>>

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
- name: Malte Jochum \footnotemark[4] 
- name: Gaetane LeProvost \footnotemark[2] 
- name: Andreas Ostrowski \footnotemark[5] 
- name: Caterina Penone \footnotemark[4] 
- name: Nadja K. Simons \footnotemark[6] 
csl: amnat.csl
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    template: pretty.template
    keep_tex: yes
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F), Senckenberganlage 25, 60325 Frankfurt am Main, Germany}
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[5]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[6]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}

@@

1.	The assessment and analysis of functional traits has been established in all domains of ecological research, promising great advancements for biodiversity and climate research. A wealth of data is available from multiple sources and for many organism groups but is lacking standardisation due to heterogeneous data formats and data types. 
2.	To harmonize and standardise datasets, we present a general vocabulary and structure for an ecological trait-data standard for storing and sharing functional trait data.
3.	We further provide an R-package that assists in transferring own data structures into the data standard and to harmonize and merge trait datasets from heterogeneous sources. 
4.	With these decentralised tools for traitdata management, we intend to facilitate the exchange and analysis of trait data for ecological research.

@@

Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. 
Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13]. These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16]. 
Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @jennings09; @simons16; @gossner16; @moretti13; @gamez-virues15; @kuhsel15;  @birkhofer14; @gravel16; @lajoie15], as well as macro-patterns of biodiversity at the global level [@diaz13; @diaz16; @reich99; @reich97]. Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].
Most importantly, the trait framework relates key ecosystem services for human well-being back to changes in biodiversity in consequence to global change or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17].

In recent years, functional ecology has thus grown into a major field of ecological research, and that continues to integrate more and more fundamental and applied research questions. 
Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. 
However, it relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa. 
To mobilize data for this enterprise, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources. The development of semantic-web standards [@wieczorek12; @page08] and ontologies [@walls12; @garnier17], as well as web-services [@kattge11; @parr16] for trait-data management is under way. 
Nonetheless, individual researchers often neglect the benefit of centralised databases and, to cope with journals' and research funding agencies increasing demand for the publication of data without access restrictions [@swan12; @allison15], instead publish primary data on file hosting services, where no data standards are forced upon the uploaded material (e.g. Figshare, Dryad). Given this prospect, there will be no lack of data, but a lack of data standardisation.

In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from an pragmatic view of data providers and data users. We further (2.) review existing trait databases and web services, as well as initiatives for standardisation. 
We (3.) describe the common practice for structuring trait data for ecological research. 
Based on these considerations, we (4.) propose a low-threshold structure and vocabulary for describing trait-datasets compatible with existing ontologies. (5.) We provide a tool for an easy adoption of this structure and for the harmonisation of data from different sources.

@@

Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12] derived from metagenomic analysis [@torsvik02; @langille13]. Databases are at hand to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g. the KEGG orthology, @kanehisa12].

@@

A global overview of existing trait data for all taxa and trait types is difficult to obtain. As an attempt to track  databases and unstandardised datasets from independent publications, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets with their their taxonomic focus and extend of traits covered, their location on the internet and the terms of use. We invite data owners and users to register further datasets qualifying as trait data.

@@

Lists of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html, @hoehndorf16]. On the side of animal traits, a couple of initiatives published comprehensive trait ontologies, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14], the Hymenoptera Anatomy Ontology [@yoder10, http://hymao.org/], or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html, @park13]. Hosting services for  biological ontologies and terminologies are the OBO foundry (http://obofoundry.org/) or ontobee (http://www.ontobee.org/).

Relating the trait definition to methodological definitions on trait measurements are key for harmonizing trait data already at the physical process of measuring. For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in Thesauri.

While trait thesauri define traits for focal groups of organisms, they are not specifying the  formats or structures in which trait data should be stored. 
The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables. The framework employs established terms provided by the Darwin Core Standard [DWC, @wieczorek12], relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g. GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr16]. Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. 
Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].

@@

Quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig. 1*a*), by aggregation of replicated measurements on multiple entities (Fig. 1*b*) or by estimation of means or ranges for the species as reported in literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1.*d*). The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or higher-level taxon.  
This definition refines the entity-quality model (EQ) as described by @garnier17, where a trait observation is phrased as 'an entity having a quality'. 
<!-- Cat: the difference [between quantitative measurements or qualitative facts] is not totally clear to me - is it the difference between field vs literature gathered data?  Flo: I see measurements as quantitative and facts as qualitative traits, which mostly aligns with field vs. literature, but not entirely. A differentiation that has rarely been specified in the trait-litereature it seems. Garnier et al 2017 just speak of "quality", which is assigned to an "entity", without specifiying it further. Nadja: I like this differentiation, to me this nicely reflects morphometric and literature traits (although its of course not exclusive). Maybe it helps if we add an example after the definition? --> <!--FDS: is this sufficiently defined now? the distinction between measured and literature data is spelled-out further below, when discussing the formats. Quantitative values can be taken from literature, though, like body sizes for example. -->

\begin{figure}[t]
\centering
\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
\caption{Types of ecological trait data assume different entities or reported quantities. a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g. reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g. a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire taxon.}

@@

Trait-datasets usually take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (a.k.a. wide-table format,  Fig. 2*a*), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as 'NA'. It may store additional information (e.g. on variation of means or literature source) in secondary columns or accompanying datasheets. 
The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon). Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also at this level of information, missing entries, e.g. about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using Bayesian approaches [@schrodt15]. However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnick16].

@@

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence). This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.  Long-table dataset purport multiple advantages for the data manipulation (e.g. filtering, sub-setting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modelling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. They further draw from a defined set of columns, which simplifies merging of datasets.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
\caption{Formats used for trait datasets. Identifiers relate the measurement to other layers of information on the specimen, the taxon or the trait measured. a) taxon-level traits are often published as a compiled species $\times$ traits matrix; b) measured data report measurements on the specimen-level, i.e. one physical occurrence of a taxon; c) a measurement long-table is a well defined and tidy data format, reporting one single measurement per row and relating it to an occurrence and taxon via unambiguous identifiers.   }

@@

With this paper, we propose a versatile, minimal data structure, as well as a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research. We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major web-services and existing standards for biodiversity data management.

@@

As described above, tidy long-table formats are the most predictable and compatible format for the core content of trait datasets and should be used for standalone uploads on low-threshold repositories.

To enable a better standardisation of nomenclatures and values, the entries of the core dataset should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names. In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08]. 
If data contain multi-layered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below). 
For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt').

@@

In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14]. Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09], the structure proposed for GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/] or EOL TraitBank [@parr16,  http://eol.org/info/cp_archives]. This also integrates a defined framework for metadata in the Ecological Metadata Language [@knb11], which should be used for describing and linking the data tables.

In some contexts, it may be safer to add such information directly as own columns of the data table to facilitate an analysis of co-factors and correlations further down the road. This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
The trade-off is data-consistency and readability *vs.* avoidance of content duplication.

Other cases, where key data columns would be placed in the same table as the core data are traits assessed on a higher level of organisation, e.g. microbial functional traits assessed at the community level taken from a soil sample. Here, location or measurement information are in the primary focus of the investigation (see vocabulary extensions below). 
A general definition, whether a column describes asset data or is part of the central dataset is ill advised. Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context.

@@

Column names and, where possible, field entries should follow the defined vocabulary listed at https://ecologicaltraitdata.github.io/TraitDataStandard/. 
Individual terms should be referenced via their global identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e. via the Application Programming Interface, API). Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

@@

It is indispensable to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits. This is achieved by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively), according to a public ontology (e.g. GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass').  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practised in TRY; @kattge11]. This redundancy of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. Keeping an additional reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
It is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].

The existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. The accessibility by software differs widely. For biodiversity data, the OBO Foundry (http://www.obofoundry.org/) or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer API services for computational access. More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.

If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. 
By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/TraitDataStandard/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for the own project context which always should be referenced in the core trait dataset. The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable webservice.

See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/TraitDataStandard/#terms-for-traitlists-a-trait-thesaurus].  We also provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).

### Metadata

Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
We define a set of terms (https://ecologicaltraitdata.github.io/TraitDataStandard/#metadata-vocabulary) that allows to relate trait data to authorship and define a bibliographic reference and license model.

@@
